$p$-adic precision

### 2014-02-20 padicprec.tex
Working on Appendix B

\begin{rem}
If $K$ has characteristic $0$ and the vector spaces $E$ and $F$ are
finite dimensional, then the $M_n$'s defined by
$$M_n = \frac 1 {|n!|} \cdot 
\sup_{\substack{1 \leq i \leq \dim E \\ \vert \underline n
\vert = n}} \,
\Big\Vert \frac{\partial^n f_i}{\partial x^{\underline n}}(0) 
\Big\Vert$$
do the job. Here $f_i$ denotes the $i$-th coordinate of $f$, the
notation $\underline n$ refers to a tuple of $(\dim F)$ nonnegative
integers and $\vert \underline n \vert$ is the sum of the coordinates
of $\underline n$.
\end{rem}x

@@

\begin{rmk}
The constant $\delta$ that appears in the above Lemma depends (up to
some multiplicative constant) on the norm that we have chosen on $T_x V$. 
However, once this norm is fixed, and assuming further that $V$ and $W$ 
are locally analytic $K$-manifolds and the mapping $f$ is locally 
analytic as well, the constant $\delta$ can be made explicit using the 
method of \S \ref{ssec:locanalytic}.
\end{rmk}

@@

In this example, we assume for simplicity that $K$ has not caracteristic 
$2$. Let $a$ and $b$ be two elements of $K$ such that $4 a^3 + 27 b^2 
\neq 0$ and let $E$ be the subset of $K^2$ consisting of the couples 
$(x,y)$ satisfying the usual equation $y^2 = x^3 + a x + b$.
Let also $\pr_x : E \to K$ (resp. $\pr_y : E \to K$) denote the map that 
takes a couple $(x,y)$ to $x$ (resp. to $y$).

We first assume that $a$ and $b$ are exactly known; typically the field 
$K$ is $\Q_p$ and $a$ and $b$ are actual integers. For each point $P_0 = 
(x_0, y_0)$ on $E$ except possibly a finite number of them, the map 
$\pr_x$ define a diffeomorphism from an open subset containing $P_0$ to 
an open subset of $K$. The same is true for $\pr_y$. Moreover around all 
$P_0 \in E$, one of these projections satisfy the above condition and
they both do generically. Hence the maps $\pr_x$ and $\pr_y$ defined an 
various open subsets define together an atlas of $E$. This turns $E$ into 
a $K$-manifold.

Let $P_0$ be a point in $E$ around which $\pr_x$ and $\pr_y$ both define 
charts. Proposition \ref{prop:independence} then tells us that a 
precision datum on $x$ determined a precision datum on $y$ and \emph{vice 
versa}. Indeed around $P_0$, we can write $y = \sqrt{x^3 + a x + b}$ 
(with the correct determination of the square root) and find the 
precision on $y$ from those on $x$ using Lemma \ref{lem:main}. We can go 
in the other direction as well by writing $x$ locally as a function of 
$y$. A precision datum at $P_0$ is then nothing but a precision datum
on the coordinate $x$ or on the coordinate $y$, keeping in mind that 
each of them determines the other. Viewing a precision datum at $P_0$ as 
a lattice in the tangent space is a nice way to make it canonical but in 
practice we can just particularize one coordinate and handle precision
only on this coordinate.

@@

Let us address now the case where $a$ and $b$ are themselves given with 
finite precision. The elliptic curve $E$ is then not fully determined. So 
we cannot consider it as a $K$-manifold and the above discussion does not 
apply readily. Nevertheless, we can always consider the submanifold of 
$K^4$ consisting of all tuples $(a,b,x,y)$ satisfying $y^2 = x^3 + a x + 
b$. The projections on the hyperplanes $a = 0$, $b = 0$, $x = 0$ and $y = 
0$ respectively define charts of this $K$-manifold. From this, we see 
that a precision datum on a point of the ``not well determined'' elliptic 
curve $E$ is a precision datum on a tuple of three variables among $a$, 
$b$, $x$ and $y$.

@@

Let $d$ and $n$ be two nonnegative integers such that $d \leq n$. The 
Grassmannian $\Grass(d,n)$ is the set of all sub-vector spaces of $K^n$ 
of dimension $d$. It defines an algebraic variety over $K$ and hence
\emph{a fortiori} a $K$-manifold.
Concretely, a vector space $V \subset K^n$ of dimension $d$ is given by 
a rectangular matrix $M \in M_{d,n}(K)$ whose rows form a basis of $V$ 
and two such matrices $M$ and $M'$ define the same vector space if there 
exists $P \in \GL_d(K)$ such that $M = P M'$. Performing row echelon, we 
find that we can always choose the above matrix $M$ in the particular
shape:
\begin{equation}
\label{eq:grass}
M = \begin{pmatrix} I_d & N \end{pmatrix} \cdot P
\end{equation}
where $I_d$ denotes the $(d \times d)$ identity matrix, $N \in M_{d, 
n-d}(K)$ and $P$ is a permutation matrix of size $n$. Moreover two 
such writings with the same $P$ necessarily coincide. Hence each 
permutation matrix $P$ defines a chart $U_P \subset \Grass(d,n)$ which
is canonically diffeomorphism to $M_{d, n-d}(K) \simeq K^{d(n-d)}$.

@@

We continue to work over a fixed ultrametric field $K$, whose absolute 
value is denoted by $|\cdot|$. We recall that usual examples of such $K$ 
are the field of $p$-adic numbers $\Q_p$ and the field of Laurent series 
$k((X))$ over a base field $k$.

We saw in the core of the article that the differential of an operation 
encodes the intrinsic loss/gain of precision when performing this 
operation. The aim of this appendix is to compute the differential of 
many usual operations on numbers, polynomials and matrices. Surprisingly, 
we observe that all differentials we will consider are rather easy to 
compute even if the underlying operation is quite involved (\emph{e.g.} 
Gr\"obner basis).

@@

A little bit more interesting is the $n$-th power map $f$ from $K$ to 
itself. Its differential $f'(x)$ is obtained by differentiate the 
equality $y = x^n$: we obtain $dy = n x^{n-1} dx$.
Hence $f'(x)$ maps to ball $B_K(r)$ (where $r$ is some positive real 
number) to $B_K(|n|{\cdot} r)$. According to Lemma \ref{lem:main}, this 
means that the behaviour of the precision depends on the absolute value 
of $n$. By the ultrametric inegality, we always have $|n| \leq 1$ but 
this inegality might be strict. It happens for instance if the 
characteristic of $K$ is divisible by $p$, in which case $n$ vanishes in 
$K$. In that case $f'(x)$ also vanishes and Lemma \ref{lem:main} does 
not apply. 
Another situation (which is more interesting) is that of $p$-adic 
numbers: let us take $K = \Q_p$ and $n = p^k$ for some integer $k$. Then
$|n| = p^{-k}$ and Lemma \ref{lem:main} reflects the well-known fact that
raising a $p$-adic number to the $p^k$-th power increases the precision 
by $k$ extra digits.

@@

For any integer $d$, let us denote by $K_{< d}[X]$ the set of 
polynomials over $K$ of degree $< d$. It is a finite dimensional vector 
space of dimension $d+1$. The affine space $X^d + K_{< d}[X]$ is then 
the set of monic polynomials over $K$ of degree $d$. We denote it by 
$K_d[X]$.

@@

Beyond sums and products (which can be treated as before), two basic 
operations involving polynomials are evaluation and interpolation.
Evaluation is modelised by the function $(P,x) \mapsto P(x)$ where
$P$ is some polynomial and $x$ some element in $K$. Differentiating
it can be done by computing
\begin{equation}
\label{eq:diffeval}
(P + dP)(x + dx) = P(x + dx) + dP(x + dx) = P(x) + P'(x) dx + dP(x)
\end{equation}
by neglecting terms of order $2$. Here $P'$ denotes the derivative of 
$P$. The differential at $P$ is then the linear map $(dP, dx) \mapsto 
P'(x) dx + dP(x)$.

@@

Let $A$ and $B$ be two polynomials $B \neq 0$. The Euclidean division of 
$A$ by $B$ is written $A = BQ + R$ with $\deg R < \deg B$. 
Differentiating the above equality we find:
$dA - dB \cdot Q = B \cdot dQ + dR$,
which implies that $dQ$ and $dR$ are respectively obtained as the 
quotient and the remainder of the Euclidean division of $dA - dB \cdot 
Q$ by $B$. This gives the differential. We note that the discussion 
above extends readily to convergent series (see also 
\cite{caruso-lubicz:14a}).

@@

We fix two positive integers $n$ and $m$ with $n \geq m$. We consider the 
function $f : K_n[X] \times K_m[X] \to (K_{\leq n}[X])^3$ which sends a 
couple $(A,B)$ to the triple $(D, U, V)$ where $D$ is the \emph{monic} 
greatest common divisor of $A$ and $B$ and $U$ and $V$ are the B\'ezout 
coefficients of minimal degrees (they are those computed by extended 
Euclide algorithm).
The nonvanishing of the resultant of $A$ and $B$ defines a Zariski open 
subset $\mathcal V_0$ where the function $\gcd$ takes the constant value 
$1$. On the contrary, outside $\mathcal V_0$, $\gcd(A,B)$ is a polynomial 
of positive degree. This implies that $f$ is not continuous outside 
$\mathcal V_0$. On the contrary, on $\mathcal V_0$, the function $f$ is
differentiable, and actually locally analytic.

@@

In order to go further and differentiate $f$ outside $\mathcal V_0$, we 
define, for all integer $i$, the subset $\mathcal V_i$ of $K_n[X] \times 
K_m[X]$ as the locus where the $\gcd$ has degree $i$. The theory of 
subresultants shows that $\mathcal V_i$ is locally closed with respect to 
the Zariski topology. In particular, it defines a $K$-manifold in the 
sense of Appendix \ref{sec:manifold}. This is interesting because the 
restriction of $f$ to $\mathcal V_i$ is now differentiable. To compute
its differential, we proceed along the same lines as before: we 
differentiate the relation $AU + BV = D$ and obtain this way
$A \cdot dU + B \cdot dV - dD = dX$
with $dX = - (dA \cdot U + dB \cdot V)$. In the above relation, the
first two terms $A{\cdot}dU$ and $B{\cdot}dV$ are divisible by $D$ 
whereas the term $dD$ has degree less than $i$. 
Hence if $dX = D \cdot dQ + dR$ is the Euclidean division of $dX$ by $D$, 
we must have $\frac A D \cdot dU + \frac B D \cdot dV = dQ$ and $dD = 
-dR$. These relations, together with bounds on the degree of $U$ and $V$,
imply as before that $dU$ (resp. $dV$) are equal to the remainder in the
Euclidean division of $U{\cdot}dQ$ by $\frac B D$ (resp. of $V{\cdot}dQ$
by $\frac A D$).

@@

The lesson we may retain from this study is the following. If we have to 
compute the greatest common divisor of two polynomials $A$ and $B$ known 
with finite precision, we first need to determine what is its degree. 
However, the degree function is not continuous --- it is only upper 
semi-continuous --- and hence cannot be determined from sure from $A$ and 
$B$ (expect if they are coprime). We then need to make an extra 
hypothesis and the more natural one is arguably to assume that 
$\gcd(A,B)$ has actually the maximal degree that it can have. Once this
assumption is made, the computation is possible and one can apply Lemma 
\ref{lem:main} to determine the precision of the result.
Note that with the above convention, the $\gcd$ of $A$ and $A$ is $A$ 
itself although there exist pair of coprime polynomials in any 
neighborhood of $(A,A)$.

@@

Suppose that we are given a polynomial $P_0 \in K_d[X]$ written as a 
product $P_0 = A_0 B_0$ where $A_0$ and $B_0$ are monic and coprime. 
Hensel's lemma implies that there exists a small neighborhood $\mathcal 
U$ of $P_0$ in $K_d[X]$ such that any $P \in \mathcal U$ factors 
uniquely as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ 
and $B_0$ respectively. Thus, we can consider the map $f : P 
\mapsto (A,B)$ defined on the subset of $\mathcal U$ consisting of monic 
polynomials. We want to differentiate $f$ at $P_0$. For this, we 
differentiate the equality $P = A B$ around $P_0$, obtaining
\begin{equation}
\label{eq:difffactor}
dP = A_0 \cdot dB + B_0 \cdot dA.
\end{equation}
where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).

@@

An important special case of the previous study occurs when $A_0$ has 
degree $1$, that is $A_0(X) = X - \alpha_0$ with some $\alpha_0 \in K$. 
The map $P \mapsto A$ is then nothing but the mapping that 
follows the simple root $\alpha_0$. Of course, its differential around 
$P_0$ can be computed by the above method. Nevertheless there is a more 
direct way to obtain it in this particular case: we write $(P_0 + 
dP)(\alpha_0 + d\alpha) = 0$ and expand this. We obtain this way 
$P_0'(\alpha_0) d\alpha + dP(\alpha_0) = 0$ where $P'$ denotes the 
derivative of $P$. Since $\alpha_0$ is a simple root, $P_0'(\alpha_0)$ 
does not vanish and we find:
$$d \alpha = - \frac{dP(\alpha_0)}{P'_0(\alpha_0)}.$$

We now address the case of a multiple roots. Let $P_0$ be a monic 
polynomial of degree $d$ and $\alpha_0 \in K$ be a root of $P_0$ having 
multiplicity $m > 1$. Due to the multiplicity, it is no longer possible 
to follow to root $\alpha_0$ in a neighborhood of $P_0$. But it is if we 
restrict ourselves to polynomials which have a root of multiplicity $m$ 
close to $\alpha_0$. More precisely, let us consider the locus $\mathcal 
V_m$ consisting of monic polynomials of degree $d$ having a root of 
multiplicity at least $m$. It is a Zariski close subset of $K_d[X]$ and 
it contains by assumption the polynomial $P_0$ we started with. 
Moreover, the irreducible components of $\mathcal V_m$ that meet at 
$P_0$ are in bijection with the set of roots of $P_0$ of multiplicity 
$\geq m$. In particular $\alpha_0$ correspond to one of these 
irreducible components; let us denote it by $\mathcal V_{m,\alpha_0}$. 
The algebraic variety $\mathcal V_{m,\alpha_0}$ is a \emph{a fortioti} 
$K$-manifold. Moreover there exists a differentiable function $f$ which 
is defined on a neighborhood of $P_0$ in $\mathcal V_{m,\alpha_0}$ and 
follows the root $\alpha_0$, \emph{i.e.} $f(P_0) = \alpha_0$ and for all 
$P$ such that $f(P)$ is defined, $\alpha = f(P)$ is a root of 
multiplicity (at least) $m$ of $P$. The existence of $f$ follows from 
Hensel's Lemma applied to the factorisation $P_0(X) = (X-\alpha_0)^m 
B_0(X)$ where $B_0(X)$ is a polynomial which is coprime to $X - 
\alpha_0$. We can finally compute the differential of $f$ at $P_0$ 
(along $\mathcal V_{m,\alpha_0}$) by differentiating the equality
$P^{(m-1)}(\alpha) = 0$ where $P^{(i)}$ denotes the $i$-th derivative
of $P$. We find:
\begin{equation}
\label{eq:dalphamult}
d \alpha = - \frac{dP^{(m-1)}(\alpha_0)}{P_0^{(m)}(\alpha_0)}.
\end{equation}

The above computation has actually interesting consequences on $p$-adic 
precision. Let us illustrate this with an exemple. Consider the monic 
polynomial $P(X) = X^2 + O(p^{2N}) X + O(p^{2N})$ (where $N$ is some 
large integer) and assume that we are asking a softwate to compute one 
of its roots. What is the right answer? We remark that, if $\alpha$ is 
any $p$-adic number divisible by $p^N$, then $P(X)$ can be $X^2 - 
\alpha^2$ whose roots are $\pm \alpha$. Conversely, we can prove that 
the two roots of $P$ are necessarily divisible by $p^N$. Hence, the
right answer is certainly $O(p^N)$. Nevertheless, if we know in addition 
that $P$ has a double root, say $\alpha$, we can write $P(X) = (X - 
\alpha)^2$ and identifying the coefficient in $X$, we get $\alpha =
O(p^{2N})$ if $p > 2$ and $\alpha = O(p^{2N-1})$ if $p=2$.
This is exactly the result we get by applying Lemma \ref{lem:main} and
using Eq.~\eqref{eq:dalphamult} to simplify the differential.

This phenomenon is general: if $P$ is a polynomial over $\Q_p$ known at 
some finite precision $O(p^N)$, a root $\alpha$ of $P$ having possibly 
multiplicity $m$ (\emph{i.e.} $P(\alpha), P'(\alpha), \ldots, P^{(m-1)} 
(\alpha)$ vanish at the given precision) can be computed at precision 
roughly $O(p^{N/m})$. However, if we know in advance that $\alpha$ has 
indeed multiplicity $m$ then we can compute it at precision $O(p^{N-c})$ 
where $c$ is some constant (depending on $P$ and $\alpha$ but not on
$N$).

>>>>

\begin{rem}
If $K$ has characteristic $0$ and the vector spaces $E$ and $F$ are
finite dimensional, then the $M_n$'s defined by
$$M_n = \frac 1 {|n!|} \cdot 
\sup_{\substack{1 \leq i \leq \dim E \\ \vert \underline n
\vert = n}} \,
\Big\Vert \frac{\partial^n f_i}{\partial x^{\underline n}}(0) 
\Big\Vert$$
do the job. Here $f_i$ denotes the $i$-th coordinate of $f$, the
notation $\underline n$ refers to a tuple of $(\dim F)$ nonnegative
integers and $\vert \underline n \vert$ is the sum of the coordinates
of $\underline n$.
\end{rem}

@@

\begin{rmk}
The constant $\delta$ that appears in the lemma depends (up to
some multiplicative constant) on the norm that we have chosen on $T_x V$. 
However, once this norm is fixed, and assuming further that $V$ and $W$ 
are locally analytic $K$-manifolds and the mapping $f$ is locally 
analytic as well, the constant $\delta$ can be made explicit using the 
method of Section \ref{ssec:locanalytic}.
\end{rmk}

@@

In this example, we assume for simplicity that $K$ does not have characteristic~$2$.
Let $a$ and $b$ be two elements of $K$ such that $4 a^3 + 27 b^2 
\neq 0$ and let $E$ be the subset of $K^2$ consisting of the pairs 
$(x,y)$ satisfying the usual equation $y^2 = x^3 + a x + b$.
Let $\pr_x : E \to K$ (resp. $\pr_y : E \to K$) denote the map that 
takes a pair $(x,y)$ to $x$ (resp. to $y$).

We first assume that $a$ and $b$ lie in the subring $R$ of exact elements. For each point $P_0 = 
(x_0, y_0)$ on $E$ except possibly a finite number of them, the map 
$\pr_x$ define a diffeomorphism from an open subset containing $P_0$ to 
an open subset of $K$; the same is true for $\pr_y$. Moreover, around each 
$P_0 \in E$, at least one of these projections satisfies the above condition.
Hence the maps $\pr_x$ and $\pr_y$ define together an atlas of $E$, giving $E$ the structure of 
a $K$-manifold.

Let $P_0$ be a point in $E$ around which $\pr_x$ and $\pr_y$ both define 
charts. Proposition \ref{prop:independence} then tells us that a 
precision datum on $x$ determines a precision datum on $y$ and vice 
versa. Indeed, in a neighborhood of $P_0$ we can write $y = \sqrt{x^3 + a x + b}$ 
(for some choice of square root) and find the 
precision on $y$ from the precision on $x$ using Lemma \ref{lem:main}. We can go 
in the other direction as well by writing $x$ locally as a function of 
$y$. A precision datum at $P_0$ is then nothing but a precision datum
on the coordinate $x$ or on the coordinate $y$, keeping in mind that 
each of them determines the other. Viewing a precision datum at $P_0$ as 
a lattice in the tangent space is a nice way to make it canonical but in 
practice we can just choose one coordinate and track precision
only on this coordinate.

@@

Finally, consider the case where $a$ and $b$ are themselves given with 
finite precision and $E$ is not fully determined. So 
we cannot consider it as a $K$-manifold and the above discussion does not 
apply readily. Nevertheless, we can always consider the submanifold of 
$K^4$ consisting of all tuples $(a,b,x,y)$ satisfying $y^2 = x^3 + a x + 
b$. The projections on the hyperplanes $a = 0$, $b = 0$, $x = 0$ and $y = 
0$ respectively define charts of this $K$-manifold. From this, we see 
that a precision datum on a point of the ``not well determined'' elliptic 
curve $E$ is a precision datum on a tuple of three variables among $a$, 
$b$, $x$ and $y$.

@@

Let $d$ and $n$ be two nonnegative integers such that $d \leq n$. The 
Grassmannian $\Grass(d,n)$ is the set of all sub-vector spaces of $K^n$ 
of dimension $d$. It defines an algebraic variety over $K$ and hence
\emph{a fortiori} a $K$-manifold.
Concretely, a vector space $V \subset K^n$ of dimension $d$ is given by 
a rectangular matrix $M \in M_{d,n}(K)$ whose rows form a basis of $V$ 
and two such matrices $M$ and $M'$ define the same vector space if there 
exists $P \in \GL_d(K)$ such that $M = P M'$. Performing row echelon, we 
find that we can always choose the above matrix $M$ in the particular
shape:
\begin{equation}
\label{eq:grass}
M = \begin{pmatrix} I_d & N \end{pmatrix} \cdot P
\end{equation}
where $I_d$ denotes the $(d \times d)$ identity matrix, $N \in M_{d, 
n-d}(K)$ and $P$ is a permutation matrix of size $n$. Moreover two 
such expressions with the same $P$ necessarily coincide. Hence each 
permutation matrix $P$ defines a chart $U_P \subset \Grass(d,n)$ which
is canonically diffeomorphic to $M_{d, n-d}(K) \simeq K^{d(n-d)}$.

@@

We saw in the core of the article that the differential of an operation 
encodes the intrinsic loss/gain of precision when performing this 
operation. The aim of this appendix is to compute the differential of 
many common operations on numbers, polynomials and matrices. Surprisingly, 
we observe that all differentials we will consider are rather easy to 
compute even if the underlying operation is quite involved (\emph{e.g.} 
Gr\"obner basis).

@@

\todofor{David}{Change to account for valuation of $x$}

Slightly more interesting is the $n$-th power map $f$ from $K$ to 
itself. Its differential $f'(x)$ is found by differentiating the 
equation $y = x^n$, obtaining $dy = n x^{n-1} dx$.
Hence $f'(x)$ maps the ball $B_K(r)$ to $B_K(|n|{\cdot} r)$. According to Lemma \ref{lem:main}, this 
means that the behavior of the precision depends on the absolute value 
of $n$. By the ultrametric inequality, we always have $|n| \leq 1$ but 
this inequality might be strict. It happens for instance if the 
characteristic of $K$ is divisible by $p$, in which case $n$ vanishes in 
$K$. In that case $f'(x)$ also vanishes and Lemma \ref{lem:main} does 
not apply. 
Another situation (which is more interesting) is that of $p$-adic 
numbers: let us take $K = \Q_p$ and $n = p^k$ for some integer $k$. Then
$|n| = p^{-k}$ and Lemma \ref{lem:main} reflects the well-known fact that
raising a $p$-adic number to the $p^k$-th power increases the precision 
by $k$ extra digits.

@@

For any integer $d$, let us denote by $K_{< d}[X]$ the set of 
polynomials over $K$ of degree $< d$. It is a finite dimensional vector 
space of dimension $d$. The affine space $X^d + K_{< d}[X]$ is then 
the set of monic polynomials over $K$ of degree $d$. We denote it by 
$K_d[X]$.

@@

Beyond sums and products (which can be treated as before), two basic 
operations involving polynomials are evaluation and interpolation.
Evaluation models the function $(P,x) \mapsto P(x)$, where
$P$ is some polynomial and $x \in K$. Differentiating
it can be done by computing
\begin{equation}
\label{eq:diffeval}
(P + dP)(x + dx) = P(x + dx) + dP(x + dx) = P(x) + P'(x) dx + dP(x)
\end{equation}
by neglecting terms of order $2$. Here $P'$ denotes the derivative of 
$P$. The differential at $P$ is then the linear map $(dP, dx) \mapsto 
P'(x) dx + dP(x)$.

@@

Let $A$ and $B$ be two polynomials with $B \neq 0$. The Euclidean division of 
$A$ by $B$ is the problem of finding $Q$ and $R$ with $A = BQ + R$ and $\deg R < \deg B$. 
Differentiating the above equality we find
\[
dA - dB \cdot Q = B \cdot dQ + dR,
\]
which implies that $dQ$ and $dR$ are respectively obtained as the 
quotient and the remainder of the Euclidean division of $dA - dB \cdot 
Q$ by $B$. This gives the differential. We note that the discussion 
above extends readily to convergent series (see also 
\cite{caruso-lubicz:14a}).

@@

We fix two positive integers $n$ and $m$ with $n \geq m$. We consider the 
function $f : K_n[X] \times K_m[X] \to (K_{\leq n}[X])^3$ which sends a 
pair $(A,B)$ to the triple $(D, U, V)$ where $D$ is the \emph{monic} 
greatest common divisor of $A$ and $B$ and $U$ and $V$ are the B\'ezout 
coefficients of minimal degrees, computed by the extended 
Euclidean algorithm.
The nonvanishing of the resultant of $A$ and $B$ defines a Zariski open 
subset $\mathcal V_0$ where the function $\gcd$ takes the constant value 
$1$. On the contrary, outside $\mathcal V_0$, $\gcd(A,B)$ is a polynomial 
of positive degree.  Since $\mathcal V_0$ is dense, $f$ is not continuous outside 
$\mathcal V_0$. Conversely, the function $f$ is differentiable, and even locally analytic,
on $\mathcal V_0$.

@@

In order to differentiate $f$ outside $\mathcal V_0$, we 
define the subset $\mathcal V_i$ of $K_n[X] \times 
K_m[X]$ as the locus where the $\gcd$ has degree $i$. The theory of 
subresultants shows that $\mathcal V_i$ is locally closed with respect to 
the Zariski topology. In particular, it defines a $K$-manifold in the 
sense of Appendix \ref{sec:manifold}, and the 
restriction of $f$ to $\mathcal V_i$ is differentiable. To compute
its differential, we proceed along the same lines as before: we 
differentiate the relation $AU + BV = D$ and obtain this way
$A \cdot dU + B \cdot dV - dD = dX$
with $dX = - (dA \cdot U + dB \cdot V)$. In the above relation, the
first two terms $A{\cdot}dU$ and $B{\cdot}dV$ are divisible by $D$ 
whereas the term $dD$ has degree less than $i$. 
Hence if $dX = D \cdot dQ + dR$ is the Euclidean division of $dX$ by $D$, 
we must have $\frac A D \cdot dU + \frac B D \cdot dV = dQ$ and $dD = 
-dR$. These relations, together with bounds on the degree of $U$ and $V$,
imply as before that $dU$ (resp. $dV$) are equal to the remainder in the
Euclidean division of $U{\cdot}dQ$ by $\frac B D$ (resp. of $V{\cdot}dQ$
by $\frac A D$).

@@

The lesson we may retain from this study is the following. If we have to 
compute the greatest common divisor of two polynomials $A$ and $B$ known 
with finite precision, we first need to determine what is its degree. 
However, the degree function is not continuous --- it is only upper 
semi-continuous --- and hence cannot be determined with certainty from $A$ and 
$B$, unless the approximation to $(A,B)$ lies entirely within $\mathcal V_0$.

We therefore need to make an extra hypothesis.  The most natural hypothesis to make
is that $\gcd(A, B)$ has the maximal possible degree.  The main reason for choosing
this convention is that if the actual polynomials $A, B \in K[X]$ have a greatest common
divisor of degree $i$ then there is some precision for which the maximal degree will be
equal to $i$, whereas if they have a positive
degree common divisor then no amount of increased precision will eliminate an intersection
with $\mathcal V_0$.  A second justification is that any other choice would yield a result
with no precision, since $A$ and $B$ appear to lie in $\mathcal V_i$ at the known precision.

Once this assumption is made, the computation is possible and one can apply Lemma 
\ref{lem:main} to determine the precision of the result.
Note that with the above convention, the $\gcd$ of $A$ and $A$ is $A$ 
itself although there exist pairs of coprime polynomials in any 
neighborhood of $(A,A)$.

@@

Suppose that we are given a polynomial $P_0 \in K_d[X]$ written as a 
product $P_0 = A_0 B_0$, where $A_0$ and $B_0$ are monic and coprime. 
Hensel's lemma implies that there exists a small neighborhood $\mathcal 
U$ of $P_0$ in $K_d[X]$ such that any $P \in \mathcal U$ factors 
uniquely as $P = A B$ with $A$ and $B$ monic and close enough to $A_0$ 
and $B_0$ respectively. Thus, we can consider the map $f : P 
\mapsto (A,B)$ defined on the subset of $\mathcal U$ consisting of monic 
polynomials. We want to differentiate $f$ at $P_0$. For this, we 
differentiate the equality $P = A B$ around $P_0$, obtaining
\begin{equation}
\label{eq:difffactor}
dP = A_0 \cdot dB + B_0 \cdot dA.
\end{equation}
where $dP$, $dA$ and $dB$ have degree less than $\deg P$, $\deg A$ and 
$\deg B$ respectively. If $A_0 U_0 + B_0 V_0 = 1$ is a Bezout relation
between $A_0$ and $B_0$, it follows from Eq.~\eqref{eq:difffactor} that
$dA$ (resp. $dB$) is the remainder in the Euclidean division of $V_0
{\cdot} dP$ by $A_0$ (resp. of $U_0 {\cdot} dP$ by $B_0$).

@@

An important special case of the previous study occurs when $A_0$ has 
degree $1$, that is $A_0(X) = X - \alpha_0$ with some $\alpha_0 \in K$. 
The map $P \mapsto A$ is then nothing but the mapping that 
follows the simple root $\alpha_0$. Of course, its differential around 
$P_0$ can be computed by the above method. Nevertheless, we may
obtain it more directly by expanding the equation $(P_0 + 
dP)(\alpha_0 + d\alpha) = 0$, obtaining
\[
P_0'(\alpha_0) d\alpha + dP(\alpha_0) = 0.
\]
Since $\alpha_0$ is a simple root, $P_0'(\alpha_0)$ 
does not vanish and we find:
$$d \alpha = - \frac{dP(\alpha_0)}{P'_0(\alpha_0)}.$$

We now address the case of a multiple roots. Let $P_0$ be a monic 
polynomial of degree $d$ and $\alpha_0 \in K$ be a root of $P_0$ having 
multiplicity $m > 1$. Due to the multiplicity, it is no longer possible 
to follow the root $\alpha_0$ in a neighborhood of $P_0$. But it is if we 
restrict ourselves to polynomials which have a root of multiplicity $m$ 
close to $\alpha_0$. More precisely, let us consider the locus $\mathcal 
V_m$ consisting of monic polynomials of degree $d$ having a root of 
multiplicity at least $m$. It is a Zariski closed subset of $K_d[X]$ and
contains by assumption the polynomial $P_0$. 
Moreover, the irreducible components of $\mathcal V_m$ that meet at 
$P_0$ are in bijection with the set of roots of $P_0$ of multiplicity 
$\geq m$. In particular, $\alpha_0$ corresponds to one of these 
irreducible components; let us denote it by $\mathcal V_{m,\alpha_0}$. 
The algebraic variety $\mathcal V_{m,\alpha_0}$ is a \emph{a fortioti} 
$K$-manifold. Moreover there exists a differentiable function $f$ which 
is defined on a neighborhood of $P_0$ in $\mathcal V_{m,\alpha_0}$ and 
follows the root $\alpha_0$, \emph{i.e.} $f(P_0) = \alpha_0$ and for all 
$P$ such that $f(P)$ is defined, $\alpha = f(P)$ is a root of 
multiplicity (at least) $m$ of $P$. The existence of $f$ follows from 
Hensel's Lemma applied to the factorisation $P_0(X) = (X-\alpha_0)^m 
B_0(X)$ where $B_0(X)$ is a polynomial which is coprime to $X - 
\alpha_0$. We can finally compute the differential of $f$ at $P_0$ 
(along $\mathcal V_{m,\alpha_0}$) by differentiating the equality
$P^{(m-1)}(\alpha) = 0$ where $P^{(i)}$ denotes the $i$-th derivative
of $P$. We find:
\begin{equation}
\label{eq:dalphamult}
d \alpha = - \frac{dP^{(m-1)}(\alpha_0)}{P_0^{(m)}(\alpha_0)}.
\end{equation}

The above computation has interesting consequences for $p$-adic 
precision. For example, consider the monic 
polynomial $P(X) = X^2 + O(p^{2N}) X + O(p^{2N})$ (where $N$ is some 
large integer) and suppose that we are asking a computer to compute one 
of its roots. What is the right answer? We remark that, if $\alpha$ is 
any $p$-adic number divisible by $p^N$, then $P(X)$ can be $X^2 - 
\alpha^2$ whose roots are $\pm \alpha$. Conversely, we can prove that 
the two roots of $P$ are necessarily divisible by $p^N$. Hence, the
right answer is certainly $O(p^N)$. Nevertheless, if we know in addition 
that $P$ has a double root, say $\alpha$, we can write $P(X) = (X - 
\alpha)^2$ and identifying the coefficient in $X$, we get $\alpha =
O(p^{2N})$ if $p > 2$ and $\alpha = O(p^{2N-1})$ if $p=2$.
This is exactly the result we get by applying Lemma \ref{lem:main} and
using Eq.~\eqref{eq:dalphamult} to simplify the differential.

This phenomenon is general: if $P$ is a polynomial over $\Q_p$ known at 
some finite precision $O(p^N)$, a root $\alpha$ of $P$ having possibly 
multiplicity $m$ (\emph{i.e.} $P(\alpha), P'(\alpha), \ldots, P^{(m-1)} 
(\alpha)$ vanish at the given precision) can be computed at precision 
roughly $O(p^{N/m})$. However, if we know in advance that $\alpha$ has 
multiplicity $m$ then we can compute it at precision $O(p^{N-c})$ 
where $c$ is some constant (depending on $P$ and $\alpha$ but not on
$N$).
